A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-11 doc/popl2013/intro.tex
tweaks related to elements, locations

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
One consequence of these decisions is that, through partitioning, a
given datum may belong to multiple different regions.  For example, if
a region $R$ is disjointly partitioned three distinct ways 
and each partition allocates every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.  There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  In fact,
the Legion implementation routinely maintains multiple distinct
physical regions, each with a copy of the data in the corresponding
logical region, for performance reasons \cite{Legion12}.

@@

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual memory
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the required privileges for function calls is checked statically
and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

>>>>

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
One consequence of these decisions is that, through partitioning, a
given datum may belong to multiple different regions.  For example, if
a region $R$ is disjointly partitioned three distinct ways 
and each partition assigns every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.  There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  In fact,
the Legion implementation routinely maintains multiple distinct
physical regions, each with a copy of the data in the corresponding
logical region, for performance reasons \cite{Legion12}.

@@

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual heap
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the required privileges for function calls are checked statically
and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

### 2012-07-11 doc/popl2013/langdef.tex
added definition of program order
tweaks related to elements, locations

Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
that still illustrates all of the important issues.  Types
include booleans, integers, tuples, and pointers.  Note that pointers
are annotated with a list of regions into which they may point; in a
well-typed program, the pointer is guaranteed to point to at least one
of those regions. There is a special type for {\em colorings}, which
are used to specify partitions of regions into subregions.
Functions in Legion are named, accept one or more arguments and
return a value of some type.  A function also declares the region
access privileges and coherence it requires.  A function may be executed
within the current task or run as a separate subtask - this decision is made
by the Legion task scheduler.  Function types are
universally quantified over all region names appearing in the type.
The final type is a region relationship, an instance of which captures a value and
one or more regions satisfying associated constraints.  A region relation instance
can be written into the heap and later read and
unpacked, giving new local names to the regions contained in
the instance.

@@

Figure~\ref{fig:validinterleave} formally defines {\em
valid\_interleave}, which we briefly explain.  Consider
traces $\oton{E}$.  The {\em exclusive location set} $L_{excl}$ is the
set of locations used in an operation with coherence mode $excl$ in
some trace $E_i$, excepting any locations in the clobber set $C$.
Similarly, the {\em atomic location set} $L_{atomic}$ is the set of
locations in an operation with coherence mode $atomic$, but no
operations with coherence mode $excl$, again excepting any location in
the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all
reads of locations in $L_2$, and any read of a location $l \in L_1$
that comes after a write to $l$, return the value most recently written
to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not 
checked, but are defered to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
test verifies every read in $E'$ with a location in $L_1$ 
is coherent if operations are performed in 
program order and that the store that results from applying $E'$
matches the store resulting from applying the
constituent traces in program order for all locations in $L_1$.  For
the atomic location set, we allow the constituent traces to be
permuted, but all locations must be checked with the same permutation.

>>>>

Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
that still illustrates all of the important issues.  Types
include booleans, integers, tuples, and pointers.  Pointers
are annotated with a list of regions - a non-null pointer must point to a 
location that is contained in at least one of the regions. There is a special
type for {\em colorings}, which
are used to specify partitions of regions into subregions.
Functions in Legion are named, accept one or more arguments and
return a value of some type.  A function also declares the region
access privileges and coherence it requires.  A function may be executed
within the current task or run as a separate subtask - this decision is made
by the Legion task scheduler.  Function types are
universally quantified over all region names appearing in the type.
The final type is a region relationship, an instance of which captures a value and
one or more regions satisfying associated constraints.  A region relation instance
can be written into the heap and later read and
unpacked, giving new local names to the regions contained in
the instance.

@@

Figure~\ref{fig:validinterleave} formally defines {\em
valid\_interleave}, which we briefly explain.  Consider
traces $\oton{E}$.  The {\em exclusive location set} $L_{excl}$ is the
set of locations used in an operation with coherence mode $excl$ in
some trace $E_i$, excepting any locations in the clobber set $C$.
Similarly, the {\em atomic location set} $L_{atomic}$ is the set of
locations in an operation with coherence mode $atomic$, but no
operations with coherence mode $excl$, again excepting any location in
the clobber set.  The $coherent(S,L_1,L_2,E)$ test checks that all
reads of locations in $L_2$, and any read of a location $l \in L_1$
that comes after a write to $l$, return the value most recently written
to the store.  (If $L_2 = \emptyset$ initially, atomic reads without a preceding write are not 
checked, but are defered to a point where traces that might be ordered before $E$ are included.)  The $seq\_equiv$
test verifies every read in $E'$ with a location in $L_1$ 
is coherent if operations are performed in 
program order and that the store that results from applying $E'$
matches the store resulting from applying the
constituent traces in program order for all locations in $L_1$.  Program order
is implicitly defined in the operational semantics, and is equivalent to a 
postorder traversal of an expression's tree.  For
the atomic location set, we allow the constituent traces to be
permuted, but all locations must be checked with the same permutation.
